Universal algebra and the \texorpdfstring{$ \lambda $}{lambda}-calculus

### 2024-06-26 report/src/algebraic-structures.tex
Write about the original representation theorem, write about cartesian closedness when L has both beta and eta equality

@@

\subsection{The $ \lambda $-calculus operations}\label{subsec:lambda-calculus-operations}
Now, for a $ \lambda $-theory $ L $, we have variables $ x_{n, i} : L_n $ and $ \lambda $-abstraction $ f \mapsto \lambda(f) $. We will sometimes denote $ \lambda(f) $ as $ \lambda x_{n + 1}, f $ for $ f: L_{n + 1} $. Now, consider the element $ \rho(x_{1, 1}) : L_2 $.
\begin{definition}
  Using the substitution, we have binary operations on the $ L_n $, sending $ (f, g) : L_n \times L_n $ to $ \rho(x_{1, 1}) \bullet (f, g) : L_n $.
\end{definition}
We will denote $ \rho(x_{1, 1}) \bullet (f, g) $ as $ f g $, and this gives us our application operation, so we can interpret all three operations of the $ \lambda $-calculus in $ L $.

\begin{remark}
  Note that if $ L $ has $ \beta $-equality, we have
  \[ \lambda(f) g = \rho(\lambda(f)) \bullet (x_1, \dots, x_n, g) = f \bullet (x_1, \dots, x_n, g), \]
  so we have the usual $ \beta $ equality. In the same way $ \eta $-equality in $ L $ gives
  \[ \lambda(\iota_{n, 1}(f) x_{n + 1}) = \lambda(\rho(f)) = f. \]
\end{remark}
\begin{remark}
  Also that for $ f: L_n $,
  \[ \rho(f) = \rho(x_1 \bullet f) = \iota_{n, 1}(f) x_{n + 1}. \]
  Therefore, we can also think of the application as being the primary operation, from which we derive $ \rho $. In the same way, we have
  \[ \rho^m(f) = \iota_{n, m}(f) x_{n + 1} \dots x_{n + m}. \]
\end{remark}
\begin{remark}
  Note that for $ f, g : L_n $,
  \[
    \rho(f) \bullet (x_1, \dots, x_n, g)
    = \rho(x_1) \bullet (\iota_{n, 1}(f), x_{n + 1}) \bullet (x_1, \dots, x_n, g)
    = \rho(x_1) \bullet (\iota_{n, 1}(f), g),
  \]
  so we could also define the application as $ f g = \rho(f) \bullet (x_1, \dots, x_n, g) $, although that is more complicated.
\end{remark}

>>>>

\begin{remark}
  Hyland claims that `a $ \lambda $-theory is an algebraic theory $ L $ equipped with semi-closed structure'. By `semi-closed structure', he probably means the structure of a semi cartesian closed category on the Lawvere theory associated to $ L $ (Lemma \ref{lem:lawvere-clone}). Because a Lawvere theory has finite products, we even would have a weak cartesian closed category. For more information about weak cartesian closed categories, see Appendix \ref{ch:weak-cartesian-closed-categories}.

  In the appendix, we see that a $ \lambda $-theory structure with $ \beta $-equality on an algebraic theory $ L $ gives its associated Lawvere theory $ C $ a weak cartesian closed structure with the exponential object $ 1^1 $ equal to $ 1 $. Conversely, we can give $ L $ a $ \lambda $-theory structure with $ \beta $-equality from a weak cartesian closed structure on $ C $ where $ 1 $ is a reflexive object. Note that we really need $ 1 $ to be a reflexive object to make the construction work.

  However, note that a weak cartesian closed structure on $ C $ with $ 1 $ a reflexive object contains more information than a $ \lambda $-theory structure with $ \beta $-equality on $ L $: If we have a $ \lambda $-theory structure with $ \beta $-equality on $ L $, then construct from this a weak cartesian closed category structure with $ 1^1 = 1 $ on $ C $ and derive from this again a $ \lambda $-theory structure on $ L $, we get the same $ \lambda $-theory structure that we started out with. If, on the other hand, we go from a weak cartesian closed structure where $ 1 $ is a reflexive object to a $ \lambda $-theory structure with $ \beta $-equality and then to a weak cartesian closed structure again, we might end up with a different structure than the one that we started out with. This is because a weak cartesian closed structure is not necessarily unique up to isomorphism. In particular, we construct semi-exponential objects $ m^n $ which are equal to $ m $ for all $ m, n : C $, and derive their data ($ \mathrm{ev} $ and $ \mathrm{cur} $) from just the data of $ 1^1 $ in the original weak cartesian closed structure. Therefore, there are enough ways in which the resulting weak cartesian closed structure can differ from the original one.

  Note that here, we really need a choice for the exponential objects. It is not enough to just ask for the mere existence of semi-exponential objects. This is because we need the information contained in $ \mathrm{ev}: L_2 $ and $ \mathrm{cur} : L_{n + 1} \to L_n $ to define $ \lambda $ and $ \rho $.

  Lastly, if $ L $ has both $ \beta $- and $ \eta $-equality, the weak cartesian closed structure becomes a cartesian closed structure, which is unique, so in that case, giving an algebraic theory $ L $ a $ \lambda $-theory structure with $ \beta $- and $ \eta $-equality is the same as giving its associated Lawvere theory a cartesian closed category structure.
\end{remark}

@@

\subsection{The $ \lambda $-calculus operations}\label{subsec:lambda-calculus-operations}
Now, for a $ \lambda $-theory $ L $, we have variables $ x_{n, i} : L_n $ and $ \lambda $-abstraction $ f \mapsto \lambda(f) $. We will sometimes denote $ \lambda(f) $ as $ \lambda x_{n + 1}, f $ for $ f: L_{n + 1} $. Now, consider the element $ \rho(x_{1, 1}) : L_2 $.
\begin{definition}
  Using the substitution, we have binary operations on the $ L_n $, sending $ (f, g) : L_n \times L_n $ to $ \rho(x_{1, 1}) \bullet (f, g) : L_n $. We will denote $ \rho(x_{1, 1}) \bullet (f, g) $ as $ f g $, and this gives us our application operation.
\end{definition}
This means that we can interpret all three operations of the $ \lambda $-calculus in $ L $.

\begin{remark}
  Note that if $ L $ has $ \beta $-equality, we have
  \[ \lambda(f) g = \rho(\lambda(f)) \bullet (x_1, \dots, x_n, g) = f \bullet (x_1, \dots, x_n, g), \]
  so we have the usual $ \beta $-equality. In the same way $ \eta $-equality of $ L $ gives
  \[ \lambda(\iota_{n, 1}(f) x_{n + 1}) = \lambda(\rho(f)) = f. \]
\end{remark}
\begin{remark}
  Also note that for $ f: L_n $,
  \[ \rho(f) = \rho(x_1 \bullet f) = \iota_{n, 1}(f) x_{n + 1}. \]
  Therefore, we can also think of the application as being the primary operation, from which we derive $ \rho $. In the same way, we have
  \[ \rho^m(f) = \iota_{n, m}(f) x_{n + 1} \dots x_{n + m}. \]
\end{remark}
\begin{remark}
  Note that for $ f, g : L_n $,
  \[
    \rho(f) \bullet (x_1, \dots, x_n, g)
    = \rho(x_1) \bullet (\iota_{n, 1}(f), x_{n + 1}) \bullet (x_1, \dots, x_n, g)
    = \rho(x_1) \bullet (\iota_{n, 1}(f), g),
  \]
  so we could also define the application as $ f g = \rho(f) \bullet (x_1, \dots, x_n, g) $, although that is more complicated.
\end{remark}

### 2024-06-26 report/src/paper.tex
Write about the original representation theorem, write about cartesian closedness when L has both beta and eta equality

>>>>

\begin{remark}
  Actually, in this thesis we stumbled upon another category with a reflexive object that exhibits $ L $ as its endomorphism theory, provided that $ L $ has $ \beta $-equality: In Appendix \ref{ch:weak-cartesian-closed-categories}, we see that the Lawvere theory $ C $ associated to $ L $ is a weak cartesian closed category, in which $ 1^1 = 1 $. It is then easy to see that the endomorphism theory that we obtain from this reflexive object $ 1 $ is exactly equal to $ L $.

  Note, however, that this is technically not an answer to Scott's original theorem, because the original formulation asks for a cartesian closed category with a reflexive object, whereas here we only have a reflexive object in a weak cartesian closed category. Even so, it is an answer to the core of the question, which is about whether we can represent any $ \lambda $-theory as the sets of morphisms $ C(U^n, U) $ for some $ U : C $.
\end{remark}

### 2024-06-26 report/src/weak-cartesian-closed.tex
Write about the original representation theorem, write about cartesian closedness when L has both beta and eta equality

We get the following as a special case, in which $ U^U $ is not only a retract of $ U $, but actually equal to $ U $.
\begin{corollary}
  Let $ L $ be an algebraic theory. Suppose that its associated Lawvere theory $ C $ is a weak cartesian closed category and suppose that for $ 1 : C $, we have $ 1^1 = 1 $. Then we can give $ L $ a $ \lambda $-theory structure with $ \beta $ equality, given by
  \[ \lambda_n(s) = \mathrm{cur}_n(s) \quad \text{and} \quad \rho(t) = \mathrm{ev} \bullet ((\iota_{n, 1}(t_i))_i + (x_{n + 1})). \]
\end{corollary}

@@

Now, we can show that giving an algebraic theory a $ \lambda $-theory structure gives its corresponding Lawvere theory (Lemma \ref{lem:lawvere-clone}) a weak cartesian closed structure.
\begin{lemma}
  Let $ L $ be a $ \lambda $-theory. Then its corresponding Lawvere theory $ C $ is a weak cartesian closed category.
\end{lemma}
\begin{proof}
  Note that by definition of a Lawvere theory, every $ n: C $ is the $ n $-fold product of $ 1 $. Therefore, binary products are given by addition on the natural numbers, and $ 0 $ is the terminal object.

@@

>>>>

We get the following as a special case, in which $ U^U $ is not only a retract of $ U $, but actually equal to $ U $.
\begin{corollary}
  Let $ L $ be an algebraic theory. Suppose that its associated Lawvere theory $ C $ is a weak cartesian closed category and suppose that for $ 1 : C $, we have $ 1^1 = 1 $. Then we can give $ L $ a $ \lambda $-theory structure with $ \beta $-equality, given by
  \[ \lambda_n(s) = \mathrm{cur}_n(s) \quad \text{and} \quad \rho(t) = \mathrm{ev} \bullet ((\iota_{n, 1}(t_i))_i + (x_{n + 1})). \]
\end{corollary}

@@

Now, we can show that giving an algebraic theory a $ \lambda $-theory structure gives its corresponding Lawvere theory (Lemma \ref{lem:lawvere-clone}) a weak cartesian closed structure.
\begin{lemma}
  Let $ L $ be a $ \lambda $-theory with $ \beta $-equality. Its corresponding Lawvere theory $ C $ is a weak cartesian closed category.
\end{lemma}
\begin{proof}
  Note that by definition of a Lawvere theory, every $ n: C $ is the $ n $-fold product of $ 1 $. Therefore, binary products are given by addition on the natural numbers, and $ 0 $ is the terminal object.

@@

\begin{remark}
  Note that if $ L $ has both $ \beta $- and $ \eta $-equality, then for $ 1^1 $,
  \[ \mathrm{cur}: C(n + 1, 1) \to C(n, 1), \]
  which is given by $ \lambda $, has an inverse given by $ \rho $. Furthermore, since $ \lambda $ and $ \rho $ respect substitution, these constitute a natural isomorphism between $ C(- + 1, 1) $ and $ C(-, 1) $, so the semi-exponential object $ 1^1 = 1 $ is in fact an exponential object. Then we have for all $ m $ and $ n $, an exponential object
  \[ m^n = (1 \times \dots \times 1)^{1 \times \dots \times 1} = 1^{1 \times \dots \times 1} \times \dots \times 1^{1 \times \dots \times 1} = 1 \times \dots \times 1 = m, \]
  so $ C $ is cartesian closed.
\end{remark}
